A Transactional Memory System for Parallel Python

### 2014-06-06 talk/dls2014/paper/paper.tex
updated benchmark numbers

For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$12.3\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$12.3\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.54\times$}
more iterations per second than a single thread.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
the others (\remi{$avg=1.9\times$} speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead from switching from GIL to STM, we see
that it is \remi{$\approx 35.5\%$}. The maximum in richards is
\remi{$63\%$}. In all benchmarks \emph{pypy-stm-nojit} beats
\emph{pypy-nojit} already on two threads despite of this overhead.
However, it rarely beats CPython's single-thread performance.  For
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

We also ran these benchmarks on Jython\footnote{version
2.7b1~\cite{webjython}}, an implementation of Python on top of the
Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
fine-grained locking for synchronisation. Even though it can use the
JVM's JIT compiler, its performance in these benchmarks is behind
PyPy's by a factor of \remi{$5-10\times$}. Because of that it does not
make sense to include it in this evaluation. In the future, we would
like to do an extensive comparison between the STM approach and
fine-grained locking. It is out of the scope of this paper to do
this thoroughly.

The results are presented in Figure~\ref{fig:performance-jit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.

@@

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of \remi{$2.5\times$} for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches \remi{$1.9\times$}.

>>>>

For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread ($12.3\%$ for all 4 threads together). While this
is not ideal, we think that $12.3\%$ are acceptable on four
threads. In terms of throughput, 4 threads have $3.54\times$
more iterations per second than a single thread.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones ($avg=2.6\times$ speedup) and a little less for
the others ($avg=2.0\times$ speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead from switching from GIL to STM, we see
that it is $\approx 43.3\%$. The maximum in richards is
$71\%$. In all benchmarks \emph{pypy-stm-nojit} beats
\emph{pypy-nojit} already on two threads despite of this overhead.
However, it rarely beats CPython's single-thread performance.  For
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

We also ran these benchmarks on Jython\footnote{version
2.7b1~\cite{webjython}}, an implementation of Python on top of the
Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
fine-grained locking for synchronisation. Even though it can use the
JVM's JIT compiler, its performance in these benchmarks is behind
PyPy's by a factor of $5-10\times$. Because of that it does not
make sense to include it in this evaluation. In the future, we would
like to do an extensive comparison between the STM approach and
fine-grained locking. It is out of the scope of this paper to do
this thoroughly.

The results are presented in Figure~\ref{fig:performance-jit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around $1-2.4\times$, and we beat GIL performance
in half of the benchmarks.

@@

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of $2.6\times$ for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches $2.0\times$.
